Application of Machine Learning to Increase the Efficiency of the Global Search Algorithm for Solving Multicriterial Problems\thanks{This research was supported by the Russian Science Foundation, project No. 21-11-00204.}

### 2023-06-10 2023/NUMTA/Full paper/NUMTA_2023_MCO.tex
1. Заключение 2. Сокрашение самоцитирования

In the class of deterministic methods for multiextremal optimization with guaranteed convergence to global optimum \cite{Gablonsky2001,Sergeyev2010,Sergeyev2006,Evtushenko2009,Evtushenko2014,Barkalov2015,Strongin2003,Sergeyev2013}, the algorithms developed in the framework of the information-statistical approach (ISA) \cite{Gergel2017_1,Gergel2018,Barkalov2014,Barkalov2018} are among the most effective. The present paper describes a new algorithm belonging to the ISA class and aimed at solving MMCO problems that accelerates the Pareto set building by the application of machine learning ideas.

The proposed method sequentially implements the concept of complexity reduction. Following this way, the initial MMCO problem is reduced to a family of single-criterion optimization subproblems by means of parameterized convolution. Then each multidimensional subproblem is transformed to an equivalent univariate (one-dimensional) optimization problem formed on the base of Peano-type mapping \cite{Evtushenko2014,Gergel2017_1,Gergel2018,Marler2004,Pardalos2017}. Further, a special global optimization algorithm designed on the ISA platform is applied for solving the univariate problems. Its computational scheme includes two accelerators that decreases significantly number of algorithm's iterations. The first one when solving current one-dimensional problem takes into account results of iterations of all the previous problems solved already. The second mechanism influences the choice of the new iteration point estimating possible closeness of its pre-image in the criteria space to the Pareto set. This estimation is carried out by the method of separation planes, which is widely used in machine learning.

@@

In our research we took weight coefficients as nodes of a uniform grid in the polyhedron (\ref{eq:06}). For assessment of the quality of Pareto set approximation obtained after solving problems (\ref{eq:08}) with mentioned weights, the HyperVolume index (HV-index) \cite{Evtushenko2014,Gergel2017_1,Gergel2018} was applied as traditional criterion of approximation accuracy.

@@

The information-statistical algorithms \cite{Gergel2017_1,Gergel2018,Barkalov2014,Barkalov2015,Barkalov2018,Strongin2003}  as deterministic methods with guaranteed convergence are among the most effective for analyzing multiextremal optimization problems. For solving multidimensional problems, these algorithms apply two approaches based on ideas of dimensionality reduction. In the framework of one approach the multidimensional problem is reduced to a family of univariate problems connected recursively (scheme of nested optimization) \cite{Grishagin2016,Zilinskas2015,Gergel2019_2}. The other approach \cite{Gergel2018,Barkalov2014,Sergeyev2013} exploits the known fact that $n$-dimensional box (1) and the closed interval $[0,1]$ of the real axis are equipotent sets and there exist unambiguous and continuous mappings of this interval onto the hyperparallelepiped (1) called \textit{Peano-type curves} or \textit{evolvents}.

@@

\section{Concludes}\label{sec:5}

>>>>

In the class of deterministic methods for multiextremal optimization with guaranteed convergence to global optimum \cite{Gablonsky2001,Sergeyev2010,Sergeyev2006,Evtushenko2014,Strongin2003,Sergeyev2013}, the algorithms developed in the framework of the information-statistical approach (ISA) \cite{Gergel2018,Barkalov2018} are among the most effective. The present paper describes a new algorithm belonging to the ISA class and aimed at solving MMCO problems that accelerates the Pareto set building by the application of machine learning ideas.

The proposed method sequentially implements the concept of complexity reduction. Following this way, the initial MMCO problem is reduced to a family of single-criterion optimization subproblems by means of parameterized convolution. Then each multidimensional subproblem is transformed to an equivalent univariate (one-dimensional) optimization problem formed on the base of Peano-type mapping \cite{Evtushenko2014,Gergel2018,Marler2004,Pardalos2017}. Further, a special global optimization algorithm designed on the ISA platform is applied for solving the univariate problems. Its computational scheme includes two accelerators that decreases significantly number of algorithm's iterations. The first one when solving current one-dimensional problem takes into account results of iterations of all the previous problems solved already. The second mechanism influences the choice of the new iteration point estimating possible closeness of its pre-image in the criteria space to the Pareto set. This estimation is carried out by the method of separation planes, which is widely used in machine learning.

@@

In our research we took weight coefficients as nodes of a uniform grid in the polyhedron (\ref{eq:06}). For assessment of the quality of Pareto set approximation obtained after solving problems (\ref{eq:08}) with mentioned weights, the HyperVolume index (HV-index) \cite{Evtushenko2014,Gergel2018} was applied as traditional criterion of approximation accuracy.

@@

The information-statistical algorithms \cite{Gergel2018,Barkalov2018,Strongin2003}  as deterministic methods with guaranteed convergence are among the most effective for analyzing multiextremal optimization problems. For solving multidimensional problems, these algorithms apply two approaches based on ideas of dimensionality reduction. In the framework of one approach the multidimensional problem is reduced to a family of univariate problems connected recursively (scheme of nested optimization) \cite{Gergel2019_2}. The other approach \cite{Gergel2018,Sergeyev2013} exploits the known fact that $n$-dimensional box (1) and the closed interval $[0,1]$ of the real axis are equipotent sets and there exist unambiguous and continuous mappings of this interval onto the hyperparallelepiped (1) called \textit{Peano-type curves} or \textit{evolvents}.

@@

\section{Conclusion}\label{sec:5}

In the paper a new numerical method for solving multidimensional multicriterial  optimization problems has been proposed for the case of multiextremal black-box partial criteria. The novelty of the method consists in application of logistic regression as a tool, widely used in machine learning, for planning iterations of the global optimization algorithm. The workability and efficiency of the method have been evaluated in a representative experiment on test classes of multiextremal problems of different dimensions. In the experiment the proposed method was compared to its version which does not utilize logistic regression and to several known evolutionary methods. All the experiments have demonstrated significant advantages of the new method over other competitors.

The further development of the research will be associated with designing parallel generalizations of the method exploiting ideas of parallel characteristical algorithms and oriented at high-performance  supercomputers.
